# Setup

This section describes how the code in the book is set up.

## Packages

Here are the packages used by the code in this book.  The last two are my own:
[tidyxl](https://nacnudus.github.io/tidyxl) and
[unpivotr](https://nacnudus.github.io/unpivotr).

```{r}
library(tibble)
library(tidyr)
library(dplyr)
library(purrr)
library(readr)
library(readxl)
library(tidyxl)
library(unpivotr)
```

You will need to install the development version of unpivotr.

```{r, eval = FALSE}
# install.packages("devtools") # If you don't already have it
devtools::install_github("nacnudus/unpivotr")
```

## Data

The examples draw from a spreadsheet of toy data, included in the
[unpivotr](https://nacnudus.github.io/unpivotr) package.  It is recommended to
[download](https://github.com/nacnudus/unpivotr/raw/master/inst/extdata/worked-examples.xlsx)
the spreadsheet and have open it in a spreadsheet application while you read the
book.

```{r}
path <- system.file("extdata", "worked-examples.xlsx", package = "unpivotr")
```

